Application Specific Arabic Morphological Analyzer for Arabic Text Mining

### 2011-02-21 tex/sarf.aclhlt.tex
some minimizations for Introduction

\begin{document}
\maketitle
\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case-based morphological 
analyzer for Arabic. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
consider the internal structure 
of an Arabic word and
compose it into several {\em morphemes}. 
A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance.
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
The efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

We hypothesize that many NLP case studies need the 
morphological analyzer to answer queries that do not need 
high accuracy at the low morphological analysis level.
The analysis at the higher levels can often compensate for 
tolerable inaccuracy at low levels. 
For example, if a query concerns proper names and the 
analyzer is considering a 
prefix that connects only to a verb,
the analyzer can answer without considering the 
rest of the word.

We find evidence to our hypothesis in~\cite{Maamouri:10}.
The addition of a new corpus to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}
required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new data challenged the existing processing and 
annotation guidelines. 
This led to refining 
SAMA~\cite{Kulick:10} with 
alternative guidelines where the NLP 
task interacted with the morphological analyzer.

The work of Habash~\shortcite{Habash:06} reports that different 
types of morphological analyses behave differently on the same case 
study. 
The analyzers may find different solutions or 
may present the solutions in different orders. 
The presented solutions and the order of presentation may not be 
appropriate for the case study. 
We find strong evidence in our results 
that a case-based morphological analyzer that allows a case-based 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

%\subsection{Sarf}
%\label{sec:intro:sarf}

{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
will be available online as an open source tool.},
 a {\em novel case-based efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by a case-based controller.
Each machine in Sarf represents a valid morphological analysis
and takes as input one letter at a time.
The case-based controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
The controller we present in Section~\ref{sec:controller}
is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
it can be a logical entity of any type.

@@

Sarf uses concatenative analysis with a novel 
{\em recursive affixes} refinement. 
An atomic affix $a$ is a standalone affix
that is not the concatenation of other affixes. 
We define an affix $f: a | a f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. 
Sarf introduces recursive affixes in order to
better retain part of speech information and enhance the 
efficiency of affix matching.

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
In this paper, we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, each 
narration into its \RL{matn} (content) and its \noArRL{sanad},
and each sanad into the 
separate narrators. Now, we can look each narrator 
up in the biography books. 
For example, 
Figure~\ref{f:exhadith} shows proper names in boxes connected
to form complex names of narrators. 
For example, 
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers %most of the analyses where we have a concentration
proper names and ignores text where
proper names occur rarely.% as long %as the text is valid.
We are also interested in discovering a words 
that relate names to each other such as \RL{bin} or ``the son of''
and words that mean narrate such as \RL{qaal} or ``said''.

With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of complex structures 
that are three level deep in hierarchy.
%We also showcase that the results of the accurate high level 
%abstraction
%results of the case-based analysis was not affected by a major
%refinement to the analyzer that increased low level analysis accuracy. 
We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a controller
guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have a case-based 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02} and define affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech (POS) tags.
\item {\bf Hadith case study:}
%We evaluated Sarf using the hadith case study. 
We were able to efficiently extract a complex three 
level hierarchical 
structure containing the chains of narrators 
with high accuracy. 
\end{enumerate}

@@

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the 
analyzer itself or use a specific set of POS tags
as their reference.
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller 
prunes these solutions.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the POS tags associated with affixes. 
%This information is highly 
%redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
finite state Automata (FSA). 
It reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a case-based
%controller. 
Beesley requires recompilation.% of the Xerox rules.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Natural language processing (NLP) uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
Sarf allows an application-specific controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf, where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
%Current morphological analyzers~\cite{Sughaiyer:04}
%use concatenative analysis when
%considering the internal structure 
%of an Arabic word and
%composing it into several {\em morphemes}. 
%A morpheme can be a {\em stem}, or an {\em affix}.
%An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
%\vocalize
%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%may have two valid morphological analyses. 
%The letter \RL{'a} may be a prefix and the word means 
%``I praise him'', or 
%%The letter \RL{'a} may also be 
%part of the stem \RL{'a.hmad} (a proper noun)
%and the word means ``his Ahmad''.
%
%\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels / diacritics. 
This greatly increases the ambiguity of Arabic text.
%(maybe reduce the couple of sentences below)
Moreover, Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

%{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
has been submitted with the paper and will be available online as an open source tool.},
 a {\em novel application-specific efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by an application-specific controller.
Each machine in Sarf takes one letter at a time as input
and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
The controller we present in Section~\ref{sec:controller}
is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
it can be a logical entity of any type.

@@

Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
{\em recursive affixes} refinement,
where an affix can be a sequence of other affixes which are concatenatively compatible.
%An atomic affix $a$ is a standalone affix
%that is not the concatenation of other affixes. 
%We define an affix $f: a | a f$ to be either
%an atomic affix $a$ or an atomic affix concatenated
%with an affix. 
Sarf introduces recursive affixes in order to
better retain part of speech information and enhance the 
efficiency of affix matching.

%Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer. In other words, the application-specific controller, has full control over and directs the morphological analysis of Sarf. % (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
%
%We hypothesize that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the low morphological analysis level.
%The analysis at the higher levels 
%(i.e. answering the application-specific query)
%can often compensate for 
%tolerable inaccuracy at lower morphological levels. 
%For example, if a query concerns proper names and the 
%analyzer is considering a 
%prefix that connects only to a verb,
%%the analyzer can answer without considering the 
%%rest of the word.
%we do not need to further analyze the rest of the word 
%to provide a negative answer.
%
%We find evidence to our hypothesis in~\cite{Maamouri:10}.
%%The addition of a new corpus to the Arabic Tree Bank 
%%(ATB)~\cite{Maamouri:04}
%%%required compliance with the DARPA GALE program 
%%%including several NLP tasks such as English translation 
%%%annotations and word-level alignment of Arabic and English. 
%%%The new data 
%%challenged the existing processing and 
%%annotation guidelines. 
%%This led to refining 
%%SAMA~\cite{Kulick:10} with 
%%alternative guidelines. 
%%This refinement was achieved only after the NLP 
%%task interacted with the morphological analyzer,
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
%In addition, 
%the work of Habash~\shortcite{Habash:06} reports that different 
%types of morphological analyses behave differently on the same %case study. 
%data set.
%The analyzers may report different solutions or 
%may present the solutions in different orders. 
%Clearly, some of 
%the presented solutions and their order of presentation may not be 
%appropriate for the %case study.
%application needs, which is an issue that Sarf tackles. % (maybe reworded)
%Moreover, 
Our results strongly support our claim that
an application-specific morphological analyzer that allows a application-specific 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

We validate our hypothesis using a case study from the Islamic 
literature, introduced later in Section~\ref{sec:islamic}.

%The rest of this paper is structured as follows. In Section~\ref{sec:related}
%we compare Sarf to related work. In Section~\ref{sec:sarf}
%we discuss Sarf and illustrate it with a running example.
%%We also explain how non-deterministic finite state Automata
%%work. 
%In Section~\ref{sec:islamic} we present the case
%study. We present our results in Section~\ref{sec:results}
%and conclude with future work in Section~\ref{sec:future}.

@@

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the 
analyzer itself or use a specific set of POS tags
as their reference.
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is application-specific in the sense that a controller 
prunes these solutions.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
%This approach has several problems. 
However, we note the following points that deserve consideration:
%(We need not to mention especially point 2 as a problem according to reviewer comments)
First, a white space delimited token may have 
more than one word.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance. % (we should stress that we mean performance in time and not accuracy although evident since some reviewers did not get)
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
However, the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow 
affixes to be defined recursively
so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the POS tags associated with affixes. 
%This information is highly 
%redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
finite state Automata (FSA). 
It reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf however constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a case-based
%controller. 
Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)
